% easychair.tex,v 3.5 2017/03/15

%\documentclass{easychair}
%\documentclass[EPiC]{easychair}
%\documentclass[EPiCempty]{easychair}
%\documentclass[debug]{easychair}
%\documentclass[verbose]{easychair}
%\documentclass[notimes]{easychair}
%\documentclass[withtimes]{easychair}
\documentclass[a4paper]{easychair}
%\documentclass[letterpaper]{easychair}

%\usepackage{doc}

% use this if you have a long article and want to create an index
% \usepackage{makeidx}

% In order to save space or manage large tables or figures in a
% landcape-like text, you can use the rotating and pdflscape
% packages. Uncomment the desired from the below.
%
% \usepackage{rotating}
% \usepackage{pdflscape}

\usepackage[T1]{fontenc}
\usepackage{microtype}
\DisableLigatures[-]{encoding = T1, family = tt* }

\usepackage{multicol}

\usepackage{doi}

\input{agda-macros}
\AgdaNoSpaceAroundCode{}

%\makeindex

%% Front Matter
%%
% Regular title as in the article class.
%
\title{Denotational Semantics of PCF in Agda \\[2ex]
\normalsize DRAFT (\today)}

% Authors are joined by \and. Their affiliations are given by \inst, which indexes
% into the list defined using \institute
%
\author{
Peter D. Mosses%\inst{1}\inst{2}
}

% Institutes for affiliations are also joined by \and,
\institute{
  Delft University of Technology, The Netherlands
  \\
  \email{p.d.mosses@tudelft.nl}
\\
   Swansea University, United Kingdom
%   \\
%   \email{p.d.mosses@swansea.ac.uk}
 }

%  \authorrunning{} has to be set for the shorter version of the authors' names;
% otherwise a warning will be rendered in the running heads. When processed by
% EasyChair, this command is mandatory: a document without \authorrunning
% will be rejected by EasyChair

\authorrunning{Peter Mosses}

% \titlerunning{} has to be set to either the main title or its shorter
% version for the running heads. When processed by
% EasyChair, this command is mandatory: a document without \titlerunning
% will be rejected by EasyChair

\titlerunning{Denotational Semantics of PCF}

\begin{document}

\maketitle

\begin{abstract}
In synthetic domain theory, all sets are predomains, domains are pointed sets, and functions are implicitly continuous.
The denotational semantics of PCF (Plotkin's version) presented here illustrates how it might look if synthetic domain theory can be implemented in Agda.
Currently, the code uses unsatisfiable postulates as a work-around, to allow Agda to type-check the definitions.

The Agda source code used to generate this document is currently available only in a private repository,
but will soon be made public.
\end{abstract}


\bigskip\hrule\bigskip


\input{PCF/All}
\clearpage
\input{PCF/Domain-Notation}
\clearpage
\input{PCF/Types}
\clearpage
\input{PCF/Constants}
\clearpage
\input{PCF/Variables}
\clearpage
\input{PCF/Environments}
\clearpage
\input{PCF/Terms}
\clearpage
\input{PCF/Checks}

\bibliographystyle{ACM-Reference-Format}
%\bibliography{PCF}

\end{document}




\begin{abstract}
  This is a progress report on ideas for improving support for Scott–Strachey denotational semantics in Agda.
  It starts by recalling relevant features of denotational semantics and Agda.
  It then 
  \end{abstract}
  
\section{Introduction}

## Domains in Denotational Semantics

In the Scott–Strachey style of denotational semantics:

- types of denotations are (Scott-)domains;
- domains are cpos with least elements, and can be defined recursively;
- denotations are defined in λ-notation, and functions are continuous;
- the isomorphisms between domains and their definitions are left implicit.

## Domains in Agda

In Agda, the [DomainTheory] modules from the [TypeTopology] library provide
well-developed support for domains.

- Domains `D` are tuples `(⟪D⟫, ⊥, _⊑_, axioms)` where:
  
  - `⟪D⟫` is a type of elements,
  - `⊥` is a distinguished element of `⟪D⟫`,
  - `_⊑_` is a partial order on `⟪D⟫`, and
  - `axioms` prove that `(⟪D⟫, _⊑_)` is a directed-complete poset (dcpo)
    with `⊥` least.

- Continuous functions `c` from domain `D` to domain `E`  are pairs
  `(f, axioms)` where:

  - `f` is an underlying function from `⟪D⟫` to `⟪E⟫`, and
  - `axioms` prove that `f` preserves limits of directed sets.

- Domains are defined recursively as bilimits of diagrams.

- Elements of domains are defined in λ-notation, where:

  - λ-abstractions need to be paired with continuity proofs,
  - applications need to select the underlying functions, and
  - the isomorphisms between domains and their definitions are explicit.

## Example: PCF

> PCF is a call-by-name simply-typed λ-calculus equipped with one or two base
> types (usually natural numbers and Booleans) and a fixed point combinator.
> In essence, PCF is the simplest lazy, purely functional programming language.
> ([PLS Lab])

PCF and its denotational semantics were originally defined by Dana Scott in 1969
([Scott 1993]) including combinators (`S`, `K`) instead of λ-abstraction.
Gordon Plotkin subsequently defined a denotational semantics for PCF including 
λ-abstraction ([Plotkin 1977]).

A direct transcription of Plotkin's definition to Agda is available in this
repo at [PCF.Plotkin.Semantics] (see also the [generated PDF]). The code
imports various modules from the [standard Agda library v2.1], and typechecks
with Agda v2.6.4.3 using the (blatantly unsafe) workarounds and postulates in
the [Domains] module.

Many authors have defined denotational semantics for PCF ([PLS Lab]). Recently,
Tom de Jong has given a definition in Agda ([DomainTheory.ScottModelOfPCF])
based on the [DomainTheory] modules from the [TypeTopology] library. The syntax
follows ([Scott 1993]) by including combinators instead of λ-abstraction.

De Jong's semantics of PCF was extended to PCF with variables and λ-abstraction
by Brendan Hart in a final year project supervised by Martín Escardó and
De Jong ([Hart 2020]).[^1] Hart used De Bruin indices for variables.

[^1]: Both De Jong and Hart deviate slightly from Scott's original semantics
    of the predecessor function by defining it to return zero when applied to
    zero, instead of being undefined.

De Jong has also given an example of how to defined a domain recursively, see
[DomainTheory.Bilimits.Dinfinity].

The definitions of the denotations of PCF terms given by De Jong and Hart
show the notational overhead that arises in Agda when using the [DomainTheory]
modules, compared to the definitions given by Scott and Plotkin.

## Extending Agda with Scott-domains

The purpose of this repo is to experiment with extending Agda to allow
denotational semantics to be defined more straightforwardly.

The main idea is to distinguish between declarations of types that correspond
to domains and declarations of ordinary types – e.g., by introducing a universe
(hierarchy) for domains.

A domain type would implicitly be a cpo, with built-in notation for its partial
order and least element. A type that corresponds to the domain of continuous 
functions from a domain to itself would also have a least fixed-point function.

The meaning of a definition in the proposed extension of Agda might be given by
translation to standard Agda, together with import of [DomainTheory] modules
from the [TypeTopology] library. Constructing the required continuity proofs
for λ-abstractions automatically could be challenging.

An alternative approach might be to typecheck code in the extension using
Agda's support for reflection and meta-programming.

Advice and suggestions are welcome, e.g., by posting to the repo [Discussions].

Peter Mosses <p.d.mosses@tudelft.nl>

[PLS Lab]: https://www.pls-lab.org/PCF "Web page"
[Scott 1993]: https://doi.org/10.1016/0304-3975(93)90095-B "TCS paper DOI"
[Plotkin 1977]: https://doi.org/10.1016/0304-3975(77)90044-5 "TCS paper DOI"
[PCF.Plotkin.Semantics]: PCF/Plotkin/Semantics.lagda "Agda module"
[Generated PDF]: latex/PCF.pdf "PDF generated by Agda"
[Domains]: Domains.lagda "Agda module"
[standard Agda library version 2.1]: https://agda.github.io/agda-stdlib/v2.1 "Agda library"
[DomainTheory]: https://www.cs.bham.ac.uk/~mhe/TypeTopology/DomainTheory.index.html "Agda modules"
[TypeTopology]: https://www.cs.bham.ac.uk/~mhe/TypeTopology "Agda library"
[DomainTheory.ScottModelOfPCF]: https://martinescardo.github.io/TypeTopology/DomainTheory.ScottModelOfPCF.ScottModelOfPCF.html "Agda module"
[Hart 2020]: https://github.com/BrendanHart/Investigating-Properties-of-PCF "GitHub repo"
[DomainTheory.Bilimits.Dinfinity]: https://martinescardo.github.io/TypeTopology/DomainTheory.Bilimits.Dinfinity.html  "Agda module"
[Discussions]: https://github.com/pdmosses/xds-agda/discussions
\section{Agda Code}
